---
title: "Lake Michigan Water Level Forecasting"
author: "Nuka Gvilia, Veda Kilaru, Anisha BharathSingh, Thomas Harmon, Ani Baghdasaryan"
date: "5/23/2022"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#########################################
# STEP 0: Set-Up                        #
#########################################

```{r}
defaultW <- getOption("warn") 
options(warn = -1)

#library(fpp)
library(tseries)
library(ggplot2)
library(forecast)
library(TSA)
library(vars)
#library(plotly)
```


```{r}
water_level <- read.csv("water_level_data.csv")
temperature <- read.csv("temperature.csv")
precipitation <- read.csv("precipitation.csv")
```

#########################################
# STEP 1: Data pre-processing           #
#########################################

```{r}
# Water level
water_level <- water_level[1:528,]
water_level.ts <- ts(water_level$MSL, start = 1978, frequency = 12) # Convert to time series object

# Temperature
temperature <- temperature[1:528, ]
temp_transpose <- data.frame(t(temperature[, 2:13]))
temp <- data.frame(Temperature=unlist(temp_transpose, use.names = FALSE))
temperature.ts <- ts(temp, start = 1978, end = c(2021, 12), frequency = 12) # Convert to time series object

# Precipitation
precipitation <- precipitation[1:528, ]
precip_transpose <- data.frame(t(precipitation[, 2:13]))
precip <- data.frame(Precipitation=unlist(precip_transpose, use.names = FALSE))
precipitation.ts <- ts(precip, start = 1978, end = c(2021, 12), frequency = 12) # Convert to time series object

# Final dataframe
df = data.frame(water_level.ts, temperature.ts, precipitation.ts)
```


#########################################
# STEP 2: Initial data exploration      #
#########################################

```{r}
plot(water_level.ts, ylab="Feet above MSL", xlab="Year", main="Average Water Level of Lake Michigan")
plot(temperature.ts, ylab="Degrees Fahrenheit", xlab="Year", main="Average Temperature of Chicago")
plot(precipitation.ts, ylab="Inches", xlab="Year", main="Average Precipitation in Chicago")
```
```{r}
plot(decompose(water_level.ts), xlab="Year")
plot(decompose(temperature.ts), xlab="Year")
plot(decompose(precipitation.ts), xlab="Year")
```
```{r}
tsdisplay(water_level.ts, ylab="Feet above MSL", xlab="Year", main="Average Water Level of Lake Michigan")
tsdisplay(temperature.ts, ylab="Degrees Fahrenheit", xlab="Year", main="Average Temperature of Chicago")
tsdisplay(precipitation.ts, ylab="Inches", xlab="Year", main="Average Precipitation in Chicago")
```
```{r}
ggAcf(water_level.ts, lag=100)
ggAcf(temperature.ts, lag=100)
ggAcf(precipitation.ts, lag=100)
```

```{r}
ggPacf(water_level.ts, lag=100)
ggPacf(temperature.ts, lag=100)
ggPacf(precipitation.ts, lag=100)
```

## CORRELATION
```{r}
cor(df)
```


```{r}
#plot_ly(x=temperature.ts, y=water_level.ts, z=precipitation.ts, type="scatter3d", mode="markers", color=temp) %>% layout(scene = list(xaxis=list(title="Temperature"), yaxis=list(title="Water Level"), zaxis=list(title="Precipitation")))

plot(x=water_level.ts, y=temperature.ts, xlab="Average Water Level", ylab="Average Temperature", main="Water Level vs. Temperature")
plot(x=water_level.ts, y=precipitation.ts, xlab="Average Water Level", ylab="Average Precipitation", main="Water Level vs. Precipitation")
plot(x=precipitation.ts, y=temperature.ts, xlab="Average Precipitation", ylab="Average Temperature", main="Precipitation vs. Temperature")
```

#########################################
# STEP 3: Data transformations          #
#########################################

```{r}
defaultW <- getOption("warn") 
options(warn = -1)

lw <- BoxCox.lambda(water_level.ts)
lw # close to 2

lt <- BoxCox.lambda(temperature.ts)
lt # close to 2

lp <- BoxCox.lambda(precipitation.ts)
lp # close to 0 i.e. ln()
```

```{r}
tsdisplay(water_level.ts)
tsdisplay(temperature.ts)
tsdisplay(precipitation.ts)
```

## Variation for all variables does not appear to increase/decrease with the level of the series; transformation not needed. 

#########################################
# STEP 4: Stationarity                  #
#########################################

```{r}
# No differencing
plot(water_level.ts, main="Water Level: No Differencing")
kpss.test(water_level.ts) # small p-value, series is *not* level stationary
adf.test(water_level.ts) # large p-value, series is *not* level stationary

plot(temperature.ts, main="Temperature: No Differencing")
kpss.test(temperature.ts) # large p-value, series *is* level stationary
adf.test(temperature.ts) # small p-value, series *is* level stationary

plot(precipitation.ts, main="Precipitation: No Differencing")
kpss.test(precipitation.ts) # large p-value, series *is* level stationary
adf.test(precipitation.ts) # small p-value, series *is* level stationary
```

```{r}
# Try with one order of seasonal differencing first for water level
water_seadiff <- diff(water_level.ts, lag=12)
plot(water_seadiff, ylab="Seasonal Difference of Water Level", xlab="Year")
acf(water_seadiff, lag=36)
pacf(water_seadiff, lag=36)
tsdisplay(water_seadiff)
tsdisplay(diff(water_seadiff))
#checkresiduals(water_seadiff)

plot(diff(water_seadiff))
acf(diff(water_seadiff), lag=100)
pacf(diff(water_seadiff), lag=100)
#checkresiduals(diff(water_seadiff))

# Try with one order regular differencing first for water level
water_diff <- diff(water_level.ts)
plot(water_diff)
acf(water_diff, lag=36)
pacf(water_diff, lag=48)

plot(diff(water_diff, lag=12))
acf(diff(water_diff, lag=12), lag=36)
pacf(diff(water_diff, lag=12), lag=48)

tsdisplay(water_diff)
tsdisplay(diff(water_diff, lag=12))

# One order regular differencing with temperature
temp_seadiff <- diff(temperature.ts, lag=12)
tsdisplay(temp_seadiff)
tsdisplay(diff(temp_seadiff))

temp_diff <- diff(temperature.ts)
tsdisplay(temp_diff)
tsdisplay(diff(temp_diff, lag=12))

# One order regular differencing with precipitation
acf(precipitation.ts, lag.max=100)
precip_seadiff <- diff(precipitation.ts, lag=12)
tsdisplay(precip_seadiff)
tsdisplay(diff(temp_seadiff))

ndiffs(temperature.ts)
nsdiffs(temperature.ts)

ndiffs(precipitation.ts)
nsdiffs(precipitation.ts)

plot(diff(water_level.ts))
tsdisplay(water_seadiff)
kpss.test(water_seadiff) # large p-value, series is level stationary
adf.test(water_seadiff) # small p-value, series is level stationary
ndiffs(water_level.ts) # 1 order of differencing required
nsdiffs(water_level.ts) # 1 order of seasonal differencing required
ndiffs(water_seadiff)
nsdiffs(water_diff)

# First order differencing
plot(diff(water_level.ts), main="Water Level: 1st Order Differencing")
kpss.test(diff(water_level.ts)) # large p-value, series is level stationary
adf.test(diff(water_level.ts)) # small p-value, series is level stationary

temp_seadiff <- diff(temperature.ts, lag=12)
tsdisplay(temp_seadiff)
kpss.test(temp_seadiff) # large p-value, series is level stationary
adf.test(temp_seadiff) # small p-value, series is level stationary
nsdiffs(temperature.ts) # 1

precip_seadiff <- diff(precipitation.ts, lag=12)
tsdisplay(precip_seadiff)
kpss.test(precip_seadiff) # large p-value, series is level stationary
adf.test(precip_seadiff) # small p-value, series is level stationary
nsdiffs(precipitation.ts) # 0
```

#########################################
# STEP 5: Train/Test split              #
#########################################
```{r}
# train time period: 1/1978 - 12/2016 
# test time period: 1/2017 - 12/2021

water_train <- window(water_level.ts, start=c(1978,1), end=c(2016,12))
water_test <- window(water_level.ts, start=c(2017,1), end=c(2021,12))

temp_train <- window(temperature.ts, start=c(1978,1), end=c(2016,12))
temp_test <- window(temperature.ts, start=c(2017,1), end=c(2021,12))

precip_train <- window(precipitation.ts, start=c(1978,1), end=c(2016,12))
precip_test <- window(precipitation.ts, start=c(2017,1), end=c(2021,12))

df_train <-data.frame(water_train, temp_train, precip_train)
df_test <- data.frame(water_test, temp_test, precip_test)

# forecast horizon
h <- as.integer(length(water_test))
```


#########################################
# STEP 6: Model & Forecast: sARIMA      #
#########################################

```{r}
arima.seasonal <- Arima(water_level.ts, order = c(2, 1, 0), seasonal = c(0, 1, 1))
arima.seasonal
checkresiduals(arima.seasonal)
Acf(arima.seasonal$residuals, 100)
```

```{r}
arima.seasonal.2 <- Arima(water_level.ts, order = c(2, 1, 0), seasonal = c(2, 1, 0))
arima.seasonal.2
checkresiduals(arima.seasonal.2)
Acf(arima.seasonal.2$residuals, 100)
```

## auto.arima()
```{r}
arima.seasonal.3 <- auto.arima(water_level.ts, seasonal = TRUE, d = 1, D = 1)
arima.seasonal.3
checkresiduals(arima.seasonal.3)
Acf(arima.seasonal.3$residuals, 100)
```
### Best model is ARIMA(2,1,0)(0,1,1)
## sARIMA Forecast
```{r}
arima.seasonal.forecast <- forecast(arima.seasonal, h=h)
plot(arima.seasonal.forecast, ylab="Water Level")
autoplot(water_test) + autolayer(arima.seasonal.forecast$mean, series="Forecast")
```

###########################################################
# STEP 7: Model & Forecast: regression with ARIMA errors  #
###########################################################


```{r}
xreg <- cbind(temperature.ts, precipitation.ts)
arima.1 <- Arima(water_level.ts, xreg = xreg, order = c(1, 1, 0), seasonal = c(0, 1, 0))
arima.1
checkresiduals(arima.1)
acf(arima.1$residuals, 100)
Acf(arima.1$residuals, 100)
```

```{r}
arima.2 <- auto.arima(water_level.ts, xreg = xreg, D = 1, d = 1)
arima.2
acf(residuals(arima.2))
checkresiduals(arima.2)
Acf(arima.2$residuals, 100)
```
### Regression with ARIMA(2,1,1)(2,1,0) is the best model.

## Forecast
```{r}
## Forecast regressions with seasonal naive method
temperature.fcst <- snaive(temperature.ts, 60)$mean
precipitation.fcst <- snaive(precipitation.ts, 60)$mean
x_reg.fcst <- cbind(temperature.fcst, precipitation.fcst)
arima.reg.forecast <- forecast(arima.2, xreg = x_reg.fcst, h=h)
plot(arima.reg.forecast, ylab="Water Level")
```

###################################################
# STEP 7: Model & Forecast: Spectral Analysis     #
###################################################

```{r}
# Find best k
AIC = rep(NA, 6)
K = c()

for (k in 1:6){
  harmonics <- fourier(water_level.ts, K = k)
  fit <- auto.arima(water_level.ts, xreg = harmonics, seasonal = FALSE)
  aic <- AIC(fit)
  AIC[k] <- aic
}

AIC # Best k is 3

harmonics <- fourier(water_level.ts, K = 3)
spectral.1 <- auto.arima(water_level.ts, xreg = harmonics, seasonal = FALSE)
spectral.1
checkresiduals(spectral.1)
```

## Forecast
```{r}
spectral.forecast <- forecast(spectral.1, xreg = fourier(water_level.ts, 3, 48))
plot(spectral.forecast, ylab="Water Level")
```

#######################################
# STEP 7: Model & Forecast: VAR       #
#######################################

```{r}
var.1<- VAR(y=data.frame(water_level.ts, temperature.ts, precipitation.ts), p=3, type='both', season=12)
summary(var.1)
preds<- predict(var.1, n.ahead=60)

# Temperature
temp_pred<- ts(preds$fcst$temp[,1], start=c(2022,1), fr=12)
ts.plot(cbind(temperature.ts, temp_pred), lty=1:2, main="Temperature Prediction", ylab="Temperature") 
ts.plot(temp_pred, main="Temperature Prediction", ylab="Temperature")

# Precipitation
precip_pred<- ts(preds$fcst$precip[,1], start=c(2022,1), fr=12)
ts.plot(cbind(precipitation.ts, precip_pred), lty=1:2, main="Precipitation Prediction", ylab="Precipitation") 
ts.plot(precip_pred, main="Precipitation Prediction", ylab="Precipitation")

# Water Level
water_pred<- ts(preds$fcst$water_level.ts[,1], start=c(2022,1), fr=12)
ts.plot(cbind(water_level.ts, water_pred), lty=1:2, main="Water Level Prediction", ylab="Water Level") 
ts.plot(water_pred, main="Water Level Prediction", ylab="Water")
ts.plot(residuals(var.1), col=1:3)
acf(residuals(var.1), 50)
pacf(residuals(var.1), 50)
```



#######################################
# STEP 8: Model Testing               #
#######################################


## Cross-Valitation with rolling window

```{r}
k <- 408 # minimum data length for fitting a model
n <- 528 # Number of data points

p <- 12 ### Period
H <- 60 # Forecast Horizon
```

```{r}
st <- tsp(water_level.ts)[1]+(k-2)/p #  gives the start time in time units,


error.seasonal.arima.1 <- matrix(NA,n-k,H)
error.seasonal.arima.2 <- matrix(NA,n-k,H)
error.arima.1 <- matrix(NA,n-k,H)
error.arima.2 <- matrix(NA,n-k,H)
error.spectral <- matrix(NA,n-k,H)


mape.seasonal.arima.1 <- matrix(NA,n-k,H)
mape.seasonal.arima.2 <- matrix(NA,n-k,H)
mape.arima.1 <- matrix(NA,n-k,H)
mape.arima.2 <- matrix(NA,n-k,H)
mape.spectral <- matrix(NA,n-k,H)

rmse_1.seasonal.arima.1 <- matrix(NA,n-k,H)
rmse_1.seasonal.arima.2 <- matrix(NA,n-k,H)
rmse_1.arima.1 <- matrix(NA,n-k,H)
rmse_1.arima.2 <- matrix(NA,n-k,H)
rmse_1.spectral <- matrix(NA,n-k,H)
rmse_1.var <- matrix(NA,n-k,H)
rmse_1.base <- matrix(NA,n-k,H)

mape_1.seasonal.arima.1 <- matrix(NA,n-k,H)
mape_1.seasonal.arima.2 <- matrix(NA,n-k,H)
mape_1.arima.1 <- matrix(NA,n-k,H)
mape_1.arima.2 <- matrix(NA,n-k,H)
mape_1.spectral <- matrix(NA,n-k,H)
mape_1.var <- matrix(NA,n-k,H)

```

```{r}
defaultW <- getOption("warn") 
options(warn = -1)


for(i in 1:60)
{
  
  ### One Month rolling forecasting
  # Expanding Window 
  train_1 <- window(water_level.ts, end=st + i/p)  ## Window Length: k+i
  train_1.temperature <- window(temperature.ts, end=st + i/p)  ## Window Length: k+i
  train_1.precipitation <- window(precipitation.ts, end=st + i/p)  ## Window Length: k+i
  train_1.xreg <- cbind(train_1.temperature, train_1.precipitation)
  train_1.temperature.fcst <- snaive(train_1.temperature, 60)$mean
  train_1.precipitation.fcst <- naive(train_1.precipitation, 60)$mean
  train_1.xreg.fcst <- cbind(train_1.temperature.fcst, train_1.precipitation.fcst)
  train_1.harmonics <- fourier(train_1, K = 3)
  
  # Sliding Window - keep the training window of fixed length. 
  # The training set always consists of k observations.
  train_2 <- window(water_level.ts, start=st+(i-k+1)/p, end=st+i/p) ## Window Length: k
  
  # Test dataset
  test <- window(water_level.ts, start=st + (i+1)/p, end=st + (i+H)/p) ## Window Length: H
  
  fit_1.seasonal.arima.1 <- Arima(train_1, order=c(2,1,0), seasonal=list(order=c(0,1,1), period=p),
                include.drift=TRUE,  method="ML")
  fcast_1.seasonal.arima.1 <- forecast(fit_1.seasonal.arima.1, h=H)
  
  
  fit_1.arima.2 <- Arima(train_1, xreg = train_1.xreg, order=c(2,1,1), seasonal=list(order=c(2,1,0), period=p),
                include.drift=TRUE,  method="ML")
  fcast_1.arima.2 <- forecast(fit_1.arima.2, h=H, xreg = train_1.xreg.fcst) 

  fit_1.spectral <- auto.arima(train_1, xreg = train_1.harmonics, seasonal = FALSE)

  fcast_1.spectral <- forecast(fit_1.spectral, xreg = fourier(train_1, 3, H)) 
  
  fit_1.var <- VAR(y=data.frame(train_1, train_1.temperature, train_1.precipitation), p=3, type='both', season=12)
  
  fcast_1.var <- predict(fit_1.var, n.ahead = 60)
  water_pred.var<- fcast_1.var$fcst$train_1[,1]
  
  waterSnaive <- snaive(train_1, h=60)
  
  rmse_1.seasonal.arima.1[i,1:length(test)] <- (fcast_1.seasonal.arima.1[['mean']]-test)^2
  rmse_1.arima.2[i,1:length(test)] <- (fcast_1.arima.2[['mean']]-test)^2
  rmse_1.spectral[i,1:length(test)] <- (fcast_1.spectral[['mean']]-test)^2
  rmse_1.var[i,1:length(test)] <- (water_pred.var-test)^2
  rmse_1.base[i, 1:length(test)] <- accuracy(waterSnaive$mean, test)[,2] # RMSE
}
```


```{r}
dev.new(width=6, height=6,pointsize=12)
plot(1:60, colMeans(rmse_1.seasonal.arima.1,na.rm=TRUE), type="l",col=1,xlab="Iterations", ylab="RMSE", main = "Cross-Validation RMSE by Model (Rolling Window)", ylim = c(0, 15))
lines(1:60, colMeans(rmse_1.arima.2,na.rm=TRUE), type="l",col=2)
lines(1:60, colMeans(rmse_1.spectral,na.rm=TRUE), type="l",col=3)
lines(1:60, colMeans(rmse_1.var,na.rm=TRUE), type="l",col=4)
#lines(1:60, colMeans(rmse_1.base,na.rm=TRUE), type="l",col=5)
legend("topleft",legend=c("Seasonal ARIMA", 'Regression with ARIMA Errors', "Spectral Analysis", "VAR"),col=1:4,lty=1)

```

## Cross-Valitation with expanding window

```{r}
rmse_2.seasonal.arima.1 <- matrix(NA,n-k,H)
rmse_2.seasonal.arima.2 <- matrix(NA,n-k,H)
rmse_2.arima.1 <- matrix(NA,n-k,H)
rmse_2.arima.2 <- matrix(NA,n-k,H)
rmse_2.spectral <- matrix(NA,n-k,H)
rmse_2.var <- matrix(NA,n-k,H)
rmse_2.base <- matrix(NA,n-k,H)

mape_2.seasonal.arima.1 <- matrix(NA,n-k,H)
mape_2.seasonal.arima.2 <- matrix(NA,n-k,H)
mape_2.arima.1 <- matrix(NA,n-k,H)
mape_2.arima.2 <- matrix(NA,n-k,H)
mape_2.spectral <- matrix(NA,n-k,H)
mape_2.var <- matrix(NA,n-k,H)
```


```{r}
defaultW <- getOption("warn") 
options(warn = -1)


for(i in 1:60)
{
  train_2 <- window(water_level.ts, start=st+(i-k+1)/p, end=st+i/p)  ## Window Length: k+i
  train_2.temperature <- window(temperature.ts, start=st+(i-k+1)/p, end=st+i/p)  ## Window Length: k+i
  train_2.precipitation <- window(precipitation.ts, start=st+(i-k+1)/p, end=st+i/p)  ## Window Length: k+i
  train_2.xreg <- cbind(train_2.temperature, train_2.precipitation)
  train_2.temperature.fcst <- snaive(train_2.temperature, 60)$mean
  train_2.precipitation.fcst <- naive(train_2.precipitation, 60)$mean
  train_2.xreg.fcst <- cbind(train_2.temperature.fcst, train_2.precipitation.fcst)
  train_2.harmonics <- fourier(train_2, K = 3)
  
  # Sliding Window - keep the training window of fixed length. 
  # The training set always consists of k observations.
  train_2 <- window(water_level.ts, start=st+(i-k+1)/p, end=st+i/p) ## Window Length: k
  
  # Test dataset
  test <- window(water_level.ts, start=st + (i+1)/p, end=st + (i+H)/p) ## Window Length: H

  
  fit_2.seasonal.arima.1 <- Arima(train_2, order=c(2,1,0), seasonal=list(order=c(1,1,0), period=p),
                include.drift=TRUE,  method="ML")
  fcast_2.seasonal.arima.1 <- forecast(fit_2.seasonal.arima.1, h=H)

  
  fit_2.arima.2 <- Arima(train_2, xreg = train_2.xreg, order=c(2,1,0), seasonal=list(order=c(2,0,0), period=p),
                include.drift=TRUE,  method="ML")
  fcast_2.arima.2 <- forecast(fit_2.arima.2, h=H, xreg = train_2.xreg.fcst) 

  fit_2.spectral <- auto.arima(train_2, xreg = train_2.harmonics, seasonal = FALSE)

  fcast_2.spectral <- forecast(fit_2.spectral, xreg = fourier(train_2, 3, H)) 
  
  fit_2.var <- VAR(y=data.frame(train_2, train_2.temperature, train_2.precipitation), p=3, type='both', season=12)
  
  fcast_2.var <- predict(fit_2.var, n.ahead = 60)
  water_pred.var<- fcast_2.var$fcst$train_2[,1]
  
  waterSnaive <- snaive(train_2, h=60)
  
  rmse_2.seasonal.arima.1[i,1:length(test)] <- (fcast_2.seasonal.arima.1[['mean']]-test)^2
  rmse_2.arima.2[i,1:length(test)] <- (fcast_2.arima.2[['mean']]-test)^2
  rmse_2.spectral[i,1:length(test)] <- (fcast_2.spectral[['mean']]-test)^2
  rmse_2.var[i,1:length(test)] <- (water_pred.var-test)^2
  rmse_2.base[i, 1:length(test)] <- accuracy(waterSnaive$mean, test)[,2] # RMSE
}
```


```{r}
dev.new(width=6, height=6,pointsize=12)
plot(1:60, colMeans(rmse_2.seasonal.arima.1,na.rm=TRUE), type="l",col=1,xlab="Iterations", ylab="RMSE", main = "Cross-Validation RMSE by Model (Expanding Window)")
lines(1:60, colMeans(rmse_2.arima.2,na.rm=TRUE), type="l",col=2)
lines(1:60, colMeans(rmse_2.spectral,na.rm=TRUE), type="l",col=3)
lines(1:60, colMeans(rmse_2.var,na.rm=TRUE), type="l",col=4)
#lines(1:60, colMeans(rmse_2.base,na.rm=TRUE), type="l",col=5)
legend("topleft",legend=c("Seasonal ARIMA", 'Regression with ARIMA Errors (auto.arima)', "Spectral Analysis", "VAR"),col=1:4,lty=1)
```




